Boosting the Accuracy of Finite Difference Schemes via Optimal Time Step Selection and Non-Iterative Defect Correction

### 2009-05-25 papers/OTSPDE/OTSPDE.tex
OTSPDE: - checked in version of OTSPDE paper submitted to Applied Numerical Mathematics

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% NUMERICS MACROS
\def\tu{\tilde{u}}
\def\dt{\Delta t}
\def\dx{\Delta x}
\def\dy{\Delta y}
\def\dto{\dt_{opt}}

% MISC MACROS
\def\eg{\emph{e.g., }}
\def\ie{\emph{i.e., }}
\def\etal{\emph{et al.}}

@@

\title{Boosting the Accuracy of Finite Difference Schemes 
       via Optimal Time Step Selection}

\author{K. T. Chu\address{Vitamin D, Inc., Menlo Park, CA 94025}$^,$\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore} 
}

\runtitle{Optimal Time Step Selection for FD Schemes}
\runauthor{K. T. Chu}

@@

\begin{abstract}
In this article, we present a novel technique for boosting the order of 
accuracy of formally low-order finite difference schemes for time dependent 
partial differential equations by optimally selecting the time step used
to advance the numerical solution.  This technique allows us to extract as
much accuracy as possible from existing finite difference schemes with 
minimal additional effort.  Through straightforward numerical 
analysis arguments, we explain the origin of the boost in accuracy and 
estimate the computational cost of the resulting numerical 
method.  We demonstrate the utility of optimal time step (OTS) selection on a 
wide array of classical linear and semilinear PDEs in one and more space 
dimensions on both regular and irregular domains.  
\end{abstract}

@@

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical 
schemes~\cite{bruger_2005,gibou_2005,ito_2005,shukla_2005,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference schemes).  
When possible, the latter approach can be a powerful way to boost the accuracy 
of a numerical method \emph{without} introducing too much additional 
algorithmic (and programming) complexity.

Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
but relatively uncommon, approach for maximizing the accuracy of numerical 
schemes.  Often, all combinations of numerical parameters are considered 
equally good.  That is, the simplest model of the error as a function 
of the numerical parameters is assumed:  each numerical parameter contributes
independently to the error.  For some problems, however, optimizing the 
numerical parameters can significantly increase the accuracy of the 
numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed that 
choosing an appropriate value for the weight factor $\theta$ in the 
theta-method can decrease the error in finite element solutions to the 
diffusion equation.

In this paper, we present a novel technique for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs by optimally 
choosing the time step (and possibly adding a few correction terms).  
Optimal time step (OTS) selection is a simple method based on the observation 
that a carefully chosen time step can eliminate low-order terms in the 
discretization error, allowing a formally low-order method to deliver 
high-order accuracy.  It combines the well-known approach of using knowledge 
of the structure of the PDE to reduce numerical error with the less-common 
notion that an optimal choice of numerical parameters can boost the 
effectiveness of algorithmically simple and computationally low-cost numerical 
schemes.

As a technique for enhancing numerical methods, optimal time step selection 
has several desirable features.  Like any technique that leads to high-order 
methods, optimal time step selection produces schemes that greatly reduce the 
computational cost required to obtain a numerical solution of a desired 
accuracy.  However, its real power comes from the fact that it achieves
high-order accuracy while being based solely on very simple, formally 
low-order finite difference schemes.  In other words, OTS selection allows us
to extract as much accuracy as possible from an existing numerical scheme
with minimal additional work.  For instance, we will see how optimal time step 
selection makes it possible to achieve 4th-order accuracy for the diffusion 
equation using only a second-order stencil for the Laplacian and forward 
Euler time integration.  Moreover, high-order accuracy is not restricted to 
problems on simple, rectangular domains; irregular domains are straightforward 
to handle by appropriately setting the values in ghost 
cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
Optimal time step selection is even beneficial when 
using finite difference schemes to solve some nonlinear PDEs.  While there 
are certainly limitations to optimal time step selection, we will see that 
there are many problems where it is useful.

It is worth spending a few moments to draw the distinction between optimal 
time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book,shampine_2005} that are used in the 
context of the method of lines~\cite{iserles_book,gko_book}.  
While both methods reduce numerical error by controlling the time step, the 
numerical errors they control are fundamentally different.  Adaptive time 
stepping can only be used to reduce temporal discretization errors because it 
controls the errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Another important 
distinction between the two techniques is that optimal time step selection 
uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection and 
presenting the analysis that explains how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we show that optimal time step
selection is useful for problems in multiple space dimensions by applying it
the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
the ease with which high-order accuracy can be achieved for problems on 
arbitrary domains, the 2D diffusion equation is solved on both regular 
and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we end some concluding remarks.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step (OTS) Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the discretization errors for 
finite difference schemes.
In this section, we present and analyze the technique of optimal time step 
selection for finite difference schemes constructed to solve scalar evolution 
equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.

@@

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the finite difference approximation, 
explicitly keeping track of the leading-order terms.  In contrast to 
conventional error analysis, we do not sweep all of the errors under big O
notation.  Rather, we take advantage of the fact that derivation of the 
leading-order terms in the error is straightforward for finite difference
schemes through the use of Taylor series expansions.  Because each term in 
the error is proportional to a partial derivative of the solution, we can 
use the PDE to relate the leading-order terms to each other.  
If we have chosen an appropriate finite difference approximation for 
(\ref{eq:model_PDE}), which is relatively straightforward to do, this careful 
analysis yields a direct relationship between the leading-order spatial and 
temporal errors that allows them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a polynomial in the numerical 
parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
$P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with an order 
of accuracy that is \emph{higher} than the formal order of the original finite 
difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are positive constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
has a similar functional dependence on the grid spacing as a time step that 
satisfies a stability constraint.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
temporal discretization errors associated with the numerical scheme used to 
integrate the system of ODEs in time.  The first term arises because $F$ is 
only an approximation of the spatial derivative operators in the PDE.  As a 
result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
the spatial discretization at \emph{every} point in time.  For a spatial 
discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
$O(\dx^r)$, so the error accumulated during each time step due to 
spatial discretization error is $O(\dt \dx^r)$.  Because the method of 
lines is so frequently used to construct finite difference approximations for 
time dependent PDEs, we will focus solely on finite difference schemes of this 
type for the remainder of the paper.  It is straightforward to extend the 
results we obtain to more general finite difference schemes.

@@

\subsubsection*{Stability Considerations}
We have not yet considered the issue of numerical stability of finite
difference schemes when the time step is set to $\dto$.  For explicit time 
integration schemes, OTS selection is ineffective when the optimal time step 
is greater than the smallest stable time step.  In this situation, we can 
switch to an implicit or semi-implicit time integration scheme with the 
same order of accuracy for which the optimal time step is stable.  
While the need to shift to implicit or semi-implicit time integration results 
in a more complicated finite difference scheme, OTS selection is still 
beneficial because it allows us to extract more accuracy from the simplest, 
low-order implicit or semi-implicit methods.
Note that a new optimal time step may need to be calculated for the modified 
finite difference scheme.

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, the optimal time step is on the order of the maximum 
time step allowed by the stability constraint for an explicit method.  
As a result, using the optimal time step to advance the finite
difference scheme in time may appear to be overly restrictive and lead to poor 
computational performance.  However, this apparent drawback is more than 
compensated by the boost in the order of accuracy.  For example, when solving 
the 1D diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step 
(discussed further in Sections~\ref{sec:diffusion_eqn_1d} and
\ref{sec:diffusion_eqn_2d}).  
As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order finite difference stencils.  The performance gain for 
problems in higher spatial dimensions is even more impressive.  As we can see 
in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tb]
\caption{
Computational cost of various finite difference schemes for the 1D 
diffusion equation as a function of the global numerical error $e$.
For all of the schemes, the standard second-order central difference 
stencil is used to discretize the Laplacian.  
Note that the time step for the backward Euler scheme is chosen so that the
temporal error is subdominant to the spatial error and that the computation 
time for the Crank-Nicholson method assumes an $O(N) = O \left( 1/\dx\right)$ 
matrix inversion algorithm. 
}
\label{tab:comp_perf_vs_err} 
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lcccc}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$
  & {\bf Memory}
  & {\bf Compute Time}
  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
% Resubmission draft: 2009/05/25, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% NUMERICS MACROS
\def\tu{\tilde{u}}
\def\he{\hat{e}}
\def\dt{\Delta t}
\def\dx{\Delta x}
\def\dy{\Delta y}
\def\dto{\dt_{opt}}

% MISC MACROS
\def\eg{\emph{e.g. }}
\def\ie{\emph{i.e. }}
\def\etal{\emph{et al.}}

@@

\title{Boosting the Accuracy of Finite Difference Schemes via
       Optimal Time Step Selection and Non-Iterative Defect Correction}

\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
}

\runtitle{Optimal Time Step Selection and Non-Iterative Defect Correction for 
          FD Schemes}
\runauthor{K.T. Chu}

@@

\begin{abstract}
In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent 
partial differential equations by optimally selecting the time step used
to advance the numerical solution and adding defect correction terms in a
non-iterative manner.  The power of the technique is its ability to extract 
as much accuracy as possible from existing finite difference schemes with 
minimal additional effort.  Through straightforward numerical analysis 
arguments, we explain the origin of the boost in accuracy and estimate the 
computational cost of the resulting numerical method.  We demonstrate the 
utility of optimal time step (OTS) selection combined with non-iterative 
defect correction (NIDC) on several different types of finite difference
schemes for a wide array of classical linear and semilinear PDEs in one and 
more space dimensions on both regular and irregular domains.  
\end{abstract}

@@

% KEYWORDS
% optimal time step; non-iterative defect correction; finite difference 
% schemes; high-order accurate numerical methods; time dependent PDEs

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical 
schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference 
schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
powerful way to boost the accuracy of a numerical method \emph{without} 
introducing too much additional algorithmic (and programming) complexity.

Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
but relatively uncommon, approach for maximizing the accuracy of numerical 
schemes.  Often, all combinations of numerical parameters are considered 
equally good.  For some problems, however, optimizing the numerical parameters 
can significantly increase the accuracy of the numerical solution.  For 
example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
upwind forward Euler discretization of the linear advection equation 
completely eliminates the numerical error introduced by the 
scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
the order of accuracy that results from setting the time step to 
$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
time integration with the standard second-order central difference 
approximation for the Laplacian.  As a final example, Zhao showed that an 
appropriate choice of the weight factor $\theta$ in the theta-method can 
significantly decrease the error in finite element solutions to the diffusion 
equation~\cite{zhao_2006}.

In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs by optimally 
choosing the time step and, when necessary, adding defect correction terms
in a \emph{non-iterative} manner.  This technique is a systematic 
generalization of the ideas underlying some of the examples mentioned above.
Optimal time step (OTS) selection is based on the observation that a carefully 
chosen time step can simultaneously eliminate low-order terms in both the
spatial and temporal the discretization errors.  It combines the well-known 
approach of using knowledge of the structure of the PDE to reduce numerical 
error with the less-common notion that an optimal choice of numerical 
parameters can boost the effectiveness of algorithmically simple and 
computationally low-cost numerical schemes.  For some problems, OTS alone is 
sufficient to allow a formally low-order method to deliver high-order 
accuracy.  However, addition of defect correction terms is generally required 
to eliminate residual terms in the leading-order truncation error.  It is 
important to emphasize that, unlike traditional defect/deferred correction 
methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
which use defect correction terms to iteratively refine the solution at 
each time step, we take a non-iterative defect correction (NIDC) approach
that incorporates defect correction terms directly into the original finite 
difference scheme.

As a technique for enhancing numerical methods, OTS-NIDC has several desirable 
features.  Like any technique that leads to high-order methods, OTS-NIDC
produces schemes that greatly reduce the computational cost required to obtain 
a numerical solution of a desired accuracy.  However, its real power comes 
from the fact that it achieves high-order accuracy while being based solely on 
very simple, formally low-order finite difference schemes.  In other words, 
OTS-NIDC allows us to extract as much accuracy as possible from an existing 
numerical scheme with minimal additional work.  For instance, we will see how 
OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
equation in any number of space dimensions using only a second-order stencil 
for the Laplacian and explicit time integration.  
Moreover, high-order accuracy is not restricted to problems 
on simple, rectangular domains; irregular domains are straightforward to 
handle by appropriately setting the values in ghost 
cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
OTS-NIDC is even beneficial when using finite difference schemes to solve some 
nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
see that there are many problems where it is useful.

This article is organized as follows.  In the remainder of this section,
we compare OTS-NIDC to a few related numerical techniques.
In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
numerical schemes.  
In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
OTS-NIDC to problems in one space dimension by applying it to finite 
difference schemes for several classical linear and nonlinear PDEs.  
In Section~\ref{sec:applications_multidim}, we show that 
OTS-NIDC is useful for problems in multiple space dimensions by applying it
the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
the ease with which high-order accuracy can be achieved for problems on 
arbitrary domains, the 2D diffusion equation is solved on both regular 
and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size, the defect correction terms, and the 
order of accuracy.  Finally, we offer some concluding remarks in 
Section~\ref{sec:summary}.

\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
Defect/deferred correction is a well-known strategy for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs.  It usually 
takes one of two forms: 
(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
or 
(2) derivation of high-order compact schemes for the spatial derivative 
operator treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.

For iterated defect/deferred correction, the numerical solution at each 
time step is computed by iteratively adding correction terms that are 
calculated from lower accuracy iterates.  When used to improve the 
temporal order of accuracy, the iterative approach can help to ensure 
stability of the time integration scheme~\cite{kress_2006}.  Although there 
are instances where iterative defect correction has been used to eliminate 
both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
during each time step because it computes correction terms using only the 
solution at the previous time step(s) (as opposed to requiring an estimate 
of the solution at the current time).

High-order compact schemes focus on eliminating spatial discretization errors 
when deriving the semi-discrete equations.  When applied to time dependent 
PDEs, a high-order compact scheme for the spatial derivative operator is 
derived treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.  
After deriving the high-order spatial discretization, any time differencing 
scheme may be applied to discretize the time derivatives.  While this 
approach is very general, it really only provides a mechanism for reducing 
spatial discretization error.  Temporal discretization errors are completely 
controlled by the choice of time differencing scheme.  Note that because 
high-order compact schemes for the spatial derivative operators involve spatial 
derivatives of the source term, the resulting schemes are always effectively 
implicit (even if all of the spatial derivative terms are treated explicitly).

OTS-NIDC has several advantages over both of these defect correction 
techniques.  First, OTS-NIDC is generally more computationally efficient than 
both iterated defect correction and the high-order compact scheme approach
because it requires less memory and, for many problems, admits simple 
explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
does not rely solely on the addition of finite differences of solution values 
to cancel out discretization errors.  Instead, it takes advantage of 
relationships between spatial derivatives, temporal derivatives and source 
terms to eliminate the need for finite differences approximations of 
high-order derivatives of the solution.  Compared with iterated defect 
correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
the presence of finite difference discretizations of high-order temporal 
derivatives and may eliminate some of the correction terms.  Comparing to the
high-order compact scheme approach, the optimization of numerical parameters 
in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
fully discrete equations.

\subsection{Comparison with Adaptive Time Stepping Methods}
OTS selection is distinct and separate from traditional adaptive time stepping 
techniques~\cite{iserles_book,shampine_2005} that are used in the 
context of the method of lines~\cite{iserles_book,gko_book}.  
While both methods reduce numerical error by controlling the time step, the 
numerical errors they control are fundamentally different.  Adaptive time 
stepping can only be used to reduce temporal discretization errors because it 
controls the errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Another important 
distinction between the two techniques is that optimal time step selection 
uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

\section{\label{sec:OTS-NIDC}
         Theoretical Foundations of OTS Selection and NIDC}
OTS-NIDC is not by itself a method for constructing finite difference 
schemes.  Rather, it is a way to enhance to the performance of existing 
finite difference schemes by carefully choosing the time step and adding
defect correction terms to eliminate low-order numerical errors.  
The two key ideas underlying OTS-NIDC are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the relationships between terms
in the discretization error. 
In this section, we present and analyze the technique of OTS-NIDC for finite 
difference schemes constructed to solve scalar evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.  The same principles apply to finite difference schemes for 
general PDEs.

@@

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the finite difference approximation, 
explicitly keeping track of the leading-order terms.  In contrast to 
conventional error analysis, we do not sweep all of the errors under big-$O$
notation.  Rather, we take advantage of the fact that derivation of the 
leading-order terms in the error is straightforward for finite difference
schemes through the use of Taylor series expansions.  Because each term in 
the error is proportional to a partial derivative of the solution, we can 
use the PDE to relate the leading-order terms to each other.  
For many finite difference schemes, this careful analysis yields a direct 
relationship between the leading-order spatial and temporal errors that allows 
them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a simple function of the 
numerical parameters $\dx$ and $\dt$.  By selecting the numerical parameters 
so that $P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with an order 
of accuracy that is \emph{higher} than the formal order of the original finite 
difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
  \label{eq:optimal_time_step}
\eeq
Notice that the optimal time step has a similar functional dependence on the 
grid spacing as a time step that satisfies a stability constraint.

An important assumption in the above analysis is that $\alpha$ and $\beta$ 
have the same sign.  If this condition is not satisfied, (\eg finite
difference scheme for the diffusion equation based on the standard central
difference stencil for the Laplacian and backward Euler time integration),
then OTS selection is not applicable because none of the zeros of 
(\ref{eq:leading_order_error_model_PDE_MOL}) is positive.  Fortunately, 
in this situation, it is usually possible to find a closely related finite 
difference scheme which \emph{does} have a positive optimal time step.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
temporal discretization errors associated with the numerical scheme used to 
integrate the system of ODEs in time.  The first term arises because $F$ is 
only an approximation of the spatial derivative operators in the PDE.  As a 
result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
the spatial discretization at \emph{every} point in time.  For a spatial 
discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
$O(\dx^r)$, so the error accumulated during each time step due to 
spatial discretization error is $O(\dt \dx^r)$.  Because the method of 
lines is so frequently used to construct finite difference approximations for 
time dependent PDEs, we will focus primarily on finite difference schemes of 
this type for the remainder of the article.  It is straightforward to extend 
the results we obtain to more general finite difference schemes (\eg the
Lax-Wendroff scheme for the linear advection equation).

@@

\subsubsection*{Stability Considerations}
We have not yet considered the issue of numerical stability of finite
difference schemes when the time step is set to $\dto$.  For explicit time 
integration schemes, OTS selection is ineffective when the optimal time step 
is greater than the smallest stable time step.  In this situation, it is 
necessary to replace the finite difference approximation of the PDE.  
For example, if the diffusion equation is solved using a 
finite difference scheme based on forward Euler time integration and a 
discretization of the Laplacian with an isotropic leading-order truncation 
error (see Section~\ref{sec:diffusion_eqn_2d}), the optimal time step becomes 
unstable when the dimension of space is greater than 3.  To make OTS selection 
useful for diffusion problems in 4 and more space dimensions, we can switch 
to the DuFort-Frankel scheme, which is explicit and unconditionally 
stable~\cite{gko_book}.

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, the optimal time step is on the order of the maximum 
time step allowed by the stability constraint for an explicit method.  
As a result, using the optimal time step to advance the finite
difference scheme in time may appear to be overly restrictive.
However, this apparent drawback is more than 
compensated by the boost in the order of accuracy.  For example, when solving 
the diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step 
(discussed further in Sections~\ref{sec:diffusion_eqn_1d} and
\ref{sec:diffusion_eqn_2d}).  
As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order finite difference stencils.  The performance gain for 
problems in higher spatial dimensions is even more impressive.  As we can see 
in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tb]
\caption{
Computational cost of various finite difference schemes for the 1D 
diffusion equation as a function of the global numerical error $e$.
For all of the schemes, the standard second-order central difference 
stencil is used to discretize the Laplacian.  
Note that the time step for the backward Euler scheme is chosen so that the
temporal error is subdominant to the spatial error and that the computation 
time for the Crank-Nicholson method assumes an $O(N) = O \left( 1/\dx\right)$ 
matrix inversion algorithm. 
}
\label{tab:comp_perf_vs_err} 
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{lcccc}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$
  & {\bf Memory}
  & {\bf Compute Time}
  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS-NIDC  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{table}
